Computational methods in landscape ecology

### 2024-10-18 Sections/06-simulations.qmd
adds tiny changes

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware]. -->

>>>>

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware]. -->

### 2024-10-18 Sections/09-conclusion.qmd
adds tiny changes

Since it emerged in the 1980s [@With2019EssentialsLandscape] the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the review provided here that showcases current computational methods in landscape ecology will only be a snapshot in time because of the development and emergence of future analytical approaches.

>>>>

Since its emergence in the 1980s [@With2019EssentialsLandscape], landscape ecology research has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change, as well as the biodiversity crisis, are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the review provided here that showcases current computational methods in landscape ecology will only be a snapshot in time because of the development and emergence of future analytical approaches.
